---
title: "R Categorical modelling"
output: html_notebook
---

### Libraries
```{r}
library(tidyverse)
library(finalfit)
library(patchwork)
library(magrittr)
library(broom)
library(ggplot2)
library(GGally)
library(lme4)
theme_set(theme_bw())
```


### Wor?ing with categorical variables
```{r}
meldata <- boot::melanoma
meldata %>% glimpse()
meldata %>% ff_glimpse()
```
#### Recode data
```{r}
meldata <- meldata %>% 
  mutate(sex.factor =             # Make new variable  
           sex %>%                # f?om existing variable
           factor() %>%           # convert to factor
           fct_recode(            # forcats function
             "Female" = "0",      # new on left, old on right
             "Male"   = "1") %>% 
           ff_label("Sex"),     ? # Optional label for finalfit
         
         # same thing but more condensed code:
         ulcer.factor = factor(ulcer) %>% 
           fct_recode("Present" = "1",
                      "Absent"  = "0") %>% 
           ff_label("Ulcerated tumour"),
 ?       
         status.factor = factor(status) %>% 
           fct_recode("Died melanoma"       = "1",
                      "Alive"               = "2",
                      "Died - other causes" = "3") %>% 
           ff_label("Status"))
```

#### Conv?rting continuous to categorical
```{r}
# Summary of age
meldata$age %>% 
  summary()
meldata %>% 
  ggplot(aes(x = age)) + 
  geom_histogram()

# Equal intervals
meldata <- meldata %>% 
  mutate(
    age.factor = 
      age %>%
      cut(4)
  )
meldata$age?factor %>%
  summary()

# Quantiles

# meldata <- meldata %>% 
#   mutate(
#     age.factor = 
#       age %>%
#       Hmisc::cut2(g=4) # Note, cut2 comes from the Hmisc package
#   )
# meldata$age.factor %>% 
#   summary()


meldata <- meldata %>% 
  muta?e(
    age.factor = 
      age %>%
      cut(breaks = c(4,20,40,60,95), include.lowest = TRUE) %>% 
      fct_recode(
        "???20"      =  "[4,20]",
        "21 to 40" = "(20,40]",
        "41 to 60" = "(40,60]",
        ">60"      = "(60,95]"
      ) %>% 
      ff_label("Age (years)")
  )

head(meldata$age.factor)
```
#### Plotting the data
```{r}
p1 <- meldata %>% 
  ggplot(aes(? = ulcer.factor, fill = status.factor)) + 
  geom_bar() + 
  theme(legend.position = "none")

p2 <- meldata %>% 
  ggplot(aes(x = ulcer.factor, fill = status.factor)) + 
  geom_bar(position = "fill") + 
  ylab("proportion")

p1 + p2

p1 <- meldata %>% 
  g?plot(aes(x = ulcer.factor, fill = status.factor)) + 
  geom_bar(position = position_stack(reverse = TRUE)) + 
  theme(legend.position = "none")

p2 <- meldata %>% 
  ggplot(aes(x = ulcer.factor, fill = status.factor)) + 
  geom_bar(position = position_fill?reverse = TRUE)) + 
  ylab("proportion")

p1 + p2

p1 <- meldata %>% 
  ggplot(aes(x = ulcer.factor, fill=status.factor)) + 
  geom_bar(position = position_stack(reverse = TRUE)) +
  facet_grid(sex.factor ~ age.factor) + 
  theme(legend.position = "none")
?p2 <- meldata %>% 
  ggplot(aes(x = ulcer.factor, fill=status.factor)) + 
  geom_bar(position = position_fill(reverse = TRUE)) +
  facet_grid(sex.factor ~ age.factor)+ 
  theme(legend.position = "bottom")

p1 / p2
```
#### Group factor levels - fct_collaps?()
```{r}
meldata <- meldata %>%
  mutate(
    status_dss = fct_collapse(
      status.factor,
      "Alive" = c("Alive", "Died - other causes"))
  )
```
#### Change order of values in factor - fct_relevel()
```{r}
# dss - disease specific survival
meldata?status_dss %>% levels()

meldata <- meldata %>% 
  mutate(status_dss = status_dss %>%
           fct_relevel("Alive")
         )
meldata$status_dss %>% levels()
```
#### Summarizing factors
```{r}
meldata %>% 
  summary_factorlist(dependent   = "status_dss?, 
                     explanatory = "ulcer.factor")

meldata %>% 
  summary_factorlist(dependent = "status_dss", 
                     explanatory = 
                       c("ulcer.factor", "age.factor", 
                         "sex.factor", "thicknes?")
  )
```
#### Chi-squared test (independence)
```{r}
table(meldata$ulcer.factor, meldata$status_dss) # both give same result
with(meldata, table(ulcer.factor, status_dss))

meldata %$%
  table(ulcer.factor, status_dss) %>% 
  prop.table(margin = 1)     #?1: row, 2: column etc.

meldata %$%        # note $ sign here
  table(ulcer.factor, status_dss) %>% 
  chisq.test()

meldata %$%        # note $ sign here
  table(ulcer.factor, status_dss) %>% 
  chisq.test() %>% 
  tidy()
```

#### Fisher's exact test
```?r}
meldata %$%        # note $ sign here
  table(age.factor, status_dss) %>% 
  chisq.test()

meldata %$%        # note $ sign here
  table(age.factor, status_dss) %>% 
  fisher.test()
```
```{r}
meldata %>% 
  summary_factorlist(dependent   = "status_dss"? 
                     explanatory = "ulcer.factor",
                     p = TRUE)

meldata %>% 
  summary_factorlist(dependent = "status_dss", 
                     explanatory = 
                       c("ulcer.factor", "age.factor", 
                  ?      "sex.factor", "thickness"),
                     p = TRUE)

meldata %>% 
  summary_factorlist(dependent = "status_dss", 
                     explanatory = 
                       c("ulcer.factor", "age.factor", 
                         "sex.factor"? "thickness"),
                     p = TRUE,
                     p_cat = "fisher")

meldata %>% 
  summary_factorlist(dependent = "status_dss", 
                     explanatory = 
                       c("ulcer.factor", "age.factor", 
                 ?       "sex.factor", "thickness"),
                     p = TRUE,
                     p_cat = "fisher",
                     digits = 
                       c(1, 1, 4, 2), #1: mean/median, 2: SD/IQR 
                                      # 3: p-value, 4:?count percentage
                     na_include = TRUE, # include missing in results/test
                     add_dependent_label = TRUE
  )
```
#### Exercise
```{r}
meldata %>% 
  summary_factorlist(dependent   = "status.factor", 
                     e?planatory = c("ulcer.factor", "age.factor", 
                         "sex.factor", "thickness"),
                     cont='median')

meldata %>%
  count(ulcer.factor, status.factor) %>%
  group_by(status.factor) %>%
  mutate(total = sum(n)) %>%
  mutate(?ercentage = round(100*n/total, 1)) %>% 
  mutate(count_perc = paste0(n, " (", percentage, ")")) %>% 
  select(-total, -n, -percentage) %>% 
  spread(status.factor, count_perc)
```
### Logistic regression
```{r}
### Preparing the data
melanoma <- boot::mela?oma
melanoma <- melanoma %>% 
  mutate(sex.factor = factor(sex) %>%          
           fct_recode("Female" = "0",
                      "Male"   = "1") %>% 
           ff_label("Sex"),   
         
         ulcer.factor = factor(ulcer) %>% 
           fc?_recode("Present" = "1",
                      "Absent"  = "0") %>% 
           ff_label("Ulcerated tumour"),
         
         age  = ff_label(age,  "Age (years)"),
         year = ff_label(year, "Year"),
         
         status.factor = factor(status)?%>% 
           fct_recode("Died melanoma"  = "1",
                      "Alive" = "2",
                      "Died - other" = "3") %>% 
           fct_relevel("Alive") %>% 
           ff_label("Status"),
         
         t_stage.factor = 
           thi?kness %>% 
           cut(breaks = c(0, 1.0, 2.0, 4.0, 
                          max(thickness, na.rm=TRUE)),
               include.lowest = TRUE)
  )

melanoma$t_stage.factor %>% levels()
melanoma <- melanoma %>% 
  mutate(
    t_stage.factor = 
      f?t_recode(t_stage.factor,
                 "T1" = "[0,1]",
                 "T2" = "(1,2]",
                 "T3" = "(2,4]",
                 "T4" = "(4,17.4]") %>% 
      ff_label("T-stage")
  )
```
```{r}
melanoma %>% 
  ggplot(aes(x = time/365)) + 
  geo?_histogram() + 
  facet_grid(. ~ status.factor)

# 5-year mortality
melanoma <- melanoma %>% 
  mutate(
    mort_5yr = 
      if_else((time/365) < 5 & 
                (status == 1), 
              "Yes",          # then
              "No") %>%       # els?
      fct_relevel("No") %>% 
      ff_label("5-year survival")
  )

p1 <- melanoma %>% 
  ggplot(aes(x = ulcer.factor, fill = mort_5yr)) + 
  geom_bar() + 
  theme(legend.position = "none")

p2 <- melanoma %>% 
  ggplot(aes(x = ulcer.factor, fill = mort_5?r)) + 
  geom_bar(position = "fill") + 
  ylab("proportion")

p1 + p2
```

```{r}
dependent <- "ulcer.factor"
explanatory <- c("age", "sex.factor", "year", "t_stage.factor")
melanoma %>% 
  summary_factorlist(dependent, explanatory, p = TRUE,
             ?       add_dependent_label = TRUE)
```
#### Model assumptions
```{r}

#### Linearity of continuous variables to the response
melanoma %>% 
  mutate(
    mort_5yr.num = as.numeric(mort_5yr) - 1
  ) %>% 
  select(mort_5yr.num, age, year) %>% 
  pivot_longer(?ll_of(c("age", "year")), names_to = "predictors") %>% 
  ggplot(aes(x = value, y = mort_5yr.num)) + 
  geom_point(size = 0.5, alpha = 0.5) +
  geom_smooth(method = "loess") + 
  facet_wrap(~predictors, scales = "free_x")

#### Multicolinearity
explanatory ?- c("ulcer.factor", "age", "sex.factor", 
                "year", "t_stage.factor")
melanoma %>% 
  remove_labels() %>%  # ggpairs doesn't work well with labels
  ggpairs(columns = explanatory)

##### Continuous to continuous
select_explanatory <- c("age",?"year")
melanoma %>% 
  remove_labels() %>% 
  ggpairs(columns = select_explanatory)

##### Continuous to categorical
select_explanatory <- c("age", "ulcer.factor", 
                       "sex.factor", "t_stage.factor")

melanoma %>% 
  select(all_of(sele?t_explanatory)) %>% 
  pivot_longer(-age) %>% # pivots all but age into two columns: name and value
  ggplot(aes(value, age)) + 
  geom_boxplot() +
  facet_wrap(~name, scale = "free", ncol = 3) +
  coord_flip()

##### Categorical to categorical
select_expl?natory <- c("ulcer.factor", "sex.factor", "t_stage.factor")

melanoma %>% 
  select(one_of(select_explanatory)) %>% 
  pivot_longer(-sex.factor) %>% 
  ggplot(aes(value, fill = sex.factor)) + 
  geom_bar(position = "fill") +
  ylab("proportion") +
  facet_?rap(~name, scale = "free", ncol = 2) +
  coord_flip()

```

```{r}
#### Variance inflation factor
# A common rule of thumb is that if this is greater than 5-10 for any variable, then multicollinearity may exist. The model should be further explored and the?terms removed or reduced.

dependent <- "mort_5yr"
explanatory <- c("ulcer.factor", "age", "sex.factor", 
                "year", "t_stage.factor")
melanoma %>% 
  glmmulti(dependent, explanatory) %>%
  car::vif()
```

##### Fitting logistic regression
```?r}
fit1 <- glm(mort_5yr ~ ulcer.factor, data = melanoma, family = binomial)
summary(fit1)
```

```{r}
coef(fit1) %>% exp()
confint(fit1) %>% exp()
fit1 %>% 
  tidy(conf.int = TRUE, exp = TRUE)
fit1 %>% 
  glance()
```
```{r}
dependent <- "mort_5yr"
explana?ory <- "ulcer.factor"
melanoma %>% 
  finalfit(dependent, explanatory, metrics = TRUE)
```
```{r}
dependent <- "mort_5yr"
explanatory <- c("ulcer.factor", "age", "sex.factor", "t_stage.factor")
fit2 = melanoma %>% 
  finalfit(dependent, explanatory, metric? = TRUE)
fit2
```
```{r}
melanoma <- melanoma %>% 
  mutate(
    age.factor = cut(age,
                     breaks = c(0, 25, 50, 75, 100)) %>% 
      ff_label("Age (years)"))

# Add this to relevel:
# fct_relevel("(50,75]")

melanoma %>% 
  finalfit(depen?ent, c("ulcer.factor", "age.factor"), metrics = TRUE)

### quadratic term
glm(mort_5yr ~ ulcer.factor  +I(age^2) + age, 
    data = melanoma, family = binomial) %>% 
  summary()

dependent <- "mort_5yr"
explanatory <- c("ulcer.factor", "I(age^2)", "age")
m?lanoma %>% 
  finalfit(dependent, explanatory, metrics = TRUE)

```
```{r}
## Reduced model
dependent <- "mort_5yr"
explanatory <- c("ulcer.factor", "age.factor", "sex.factor", "t_stage.factor")
explanatory_multi <- c("ulcer.factor", "sex.factor", "t_stage?factor")

melanoma %>% 
  finalfit(dependent, explanatory, explanatory_multi, 
           keep_models = TRUE, metrics = TRUE)

dependent <- "mort_5yr"
explanatory <- c("ulcer.factor", "sex.factor", "t_stage.factor")
explanatory_multi <- c("ulcer.factor", "?_stage.factor")

melanoma %>% 
  finalfit(dependent, explanatory, explanatory_multi, 
           keep_models = TRUE, metrics = TRUE)

dependent <- "mort_5yr"
explanatory <- c("ulcer.factor", "t_stage.factor")
explanatory_multi <- c("ulcer.factor*t_stage.fa?tor")
melanoma %>% 
  finalfit(dependent, explanatory, explanatory_multi, 
           keep_models = TRUE, metrics = TRUE)

dependent <- "mort_5yr"
explanatory <- c("ulcer.factor", "age.factor", 
                "sex.factor", "t_stage.factor")
explanatory_m?lti <- c("ulcer.factor", "t_stage.factor")
melanoma %>% 
  finalfit(dependent, explanatory, explanatory_multi, metrics = TRUE)
```

```{r}
## odds-ratio plot
dependent <- "mort_5yr"
explanatory_multi <- c("ulcer.factor", "t_stage.factor")
melanoma %>% 
  o?_plot(dependent, explanatory_multi,
          breaks = c(0.5, 1, 2, 5, 10, 25),
          table_text_size = 3.5,
          title_text_size = 16)
```
#### Correlated groups of observations
```{r}
# Simulate random hospital identifier
set.seed(1)
melanoma <-?melanoma %>% 
    mutate(hospital_id = sample(1:4, 205, replace = TRUE))

melanoma <- melanoma %>%
    mutate(hospital_id = c(rep(1:3, 50), rep(4, 55)))

melanoma %>% 
    mutate(
    mort_5yr.num = as.numeric(mort_5yr) - 1 # Convert factor to 0 and 1
  ) ?>% 
  ggplot(aes(x = as.numeric(t_stage.factor), y = mort_5yr.num)) +
  geom_jitter(width = 0.1, height = 0.1) +
  geom_smooth(method = 'loess', se = FALSE) + 
  facet_wrap(~hospital_id) + 
  labs(x= "T-stage", y = "Mortality (5 y)")
```
```{r}
melanoma %>? 
  glmer(mort_5yr ~ t_stage.factor + (1 | hospital_id), 
        data = ., family = "binomial") %>% 
  summary()

print('**********************************')
dependent <- "mort_5yr"
explanatory <- "t_stage.factor"
random_effect <- "hospital_id" # Is the s?me as:
random_effect <- "(1 | hospital_id)"
melanoma %>% 
  finalfit(dependent, explanatory, 
           random_effect = random_effect,
           metrics = TRUE)
print('**********************************')
dependent <- "mort_5yr"
explanatory <- c("ulcer.f?ctor", "age.factor", 
                "sex.factor", "t_stage.factor")
explanatory_multi <- c("ulcer.factor", "t_stage.factor")
random_effect <- "hospital_id"
melanoma %>% 
  finalfit(dependent, explanatory, explanatory_multi, random_effect, 
           kee?_models = TRUE,
           metrics = TRUE)
```
##### Exercise
```{r}
## Recode
melanoma <- melanoma %>% 
  mutate(sex.factor = factor(sex) %>%          
           fct_recode("Female" = "0",
                      "Male"   = "1") %>% 
           ff_label("S?x"),   
         
         ulcer.factor = factor(ulcer) %>% 
           fct_recode("Present" = "1",
                      "Absent"  = "0") %>% 
           ff_label("Ulcerated tumour"),
         
         age  = ff_label(age,  "Age (years)"),
         year ? ff_label(year, "Year"),
         
         status.factor = factor(status) %>% 
           fct_recode("Died melanoma"  = "1",
                      "Alive" = "2",
                      "Died - other" = "3") %>% 
           fct_relevel("Alive") %>% 
       ?   ff_label("Status"),
         
         t_stage.factor = 
           thickness %>% 
           cut(breaks = c(0, 1.0, 2.0, 4.0, 
                          max(thickness, na.rm=TRUE)),
               include.lowest = TRUE)
  )

# Plot
p1 <- melanoma %>% 
? ggplot(aes(x = sex.factor, fill = mort_5yr)) + 
  geom_bar() + 
  theme(legend.position = "none")

p2 <- melanoma %>% 
  ggplot(aes(x = sex.factor, fill = mort_5yr)) + 
  geom_bar(position = "fill") + 
  ylab("proportion")

p1 + p2
```
```{r}
## Recode T-?tage first
melanoma <- melanoma %>% 
  mutate(
    t_stage.factor = 
      fct_recode(t_stage.factor,
                 T1 = "[0,1]",
                 T2 = "(1,2]",
                 T3 = "(2,4]",
                 T4 = "(4,17.4]") %>% 
      ff_label("T-stag?")
  )

dependent = "sex.factor"
explanatory = c("age", "t_stage.factor", "ulcer.factor")
melanoma %>% 
  summary_factorlist(dependent, explanatory, p = TRUE, na_include = TRUE,
                     cont = "median")

# Men have more T4 tumours and they are?more likely to be ulcerated. 
```
```{r}
dependent = "mort_5yr"
explanatory = c("sex.factor", "age", "t_stage.factor", "ulcer.factor")
melanoma %>% 
  finalfit(dependent, explanatory, metrics = TRUE)

# c-statistic = 0.798
# In multivariable model, male vs?female OR 1.26 (0.57-2.76, p=0.558).
# No relationship after accounting for T-stage and tumour ulceration. 
# Sex is confounded by these two variables. 
```
```{r}
dependent = "mort_5yr"
explanatory = c("sex.factor", "age", "t_stage.factor", "ulcer.factor"?
melanoma %>% 
  or_plot(dependent, explanatory)
```

